---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

```{r load_packages, include = FALSE}
library(e1071)
```

# elements

<!-- badges: start -->
[![Generic badge](https://img.shields.io/badge/Version-0.0.1-green.svg)]()
[![License: CC BY 4.0](https://img.shields.io/badge/License-LGPL 3.0-lightgrey.svg)](https://opensource.org/license/lgpl-3-0)
<!-- badges: end -->

`elements` is a R package containing **e**co**l**ogical nich**e** **m**od**e**ls for the pla**n**t**s** of Europe.

The Ecological Niche Models (ENMs) adhere to Hutchinsonian conceptualisation of the ecological niche as an n-dimensional hypervolume (Hutchinson, 1957) and are comprised of `e1071` (Meyer et al., 2024) Support Vector Machine (SVM) models.

For more information please see Marshall et al (in prep).

## Installation

You can install the development version of elements from [Github](https://github.com/NERC-CEH/elements) with:

``` r
# install.packages("remotes")
remotes::install_github("NERC-CEH/elements")
```

## Usage

### Retrieving models

Due to the total size of the `r length(elements::ModellingSpecies)` ENMs currently included in the `elements` r package
the ENMs are not exported as a .rda object. Instead they are made available through a `filehash` (Peng, 2005) database.
To access the ENMs a connection to this database must be initialised, the ENMs can then be accessed like as a list but without loading all ENMs into memory.

```{r retrieve_models}
elements::startup()

model <- OccModels[["stellaria_graminea"]]
```

```{r print_model, echo=FALSE}
print(model)
```

### Using the models

The raw ENMs retrieved using the method above can be used as regular `e1071` SVM model objects. Alternatively, the helper function `elements::predict_occ` retrieves these models using the method above, and eases access to the results.

```{r use_models}
results <- elements::predict_occ(taxon = "stellaria_graminea", predictors = elements::ExampleData, pa = "Present", dp = 2)
```

```{r print_results, echo=FALSE}
head(results)
```

### Shutting down

At the end of the analyis run `elements::shutdown` to close the connection to the filehash database.

```{r shutdown}
elements::shutdown()
```

## Inspecting models

Several datasets are available to examine the ENM model performance and aid in model interpretation.

The performance measures can be retrieved from the `elements::PerformanceMeasures` object; 
for example, below the balanced accuracy from the random holdout sample and the model tuning spatio-temporal cross-validation (Schratz et al., 2024) are displayed.

```{r performance_measures}
pm <- elements::PerformanceMeasures

pm_taxon <- pm[pm[["taxon_code"]] == "stellaria_graminea", c(1, 7, 8)]
```

```{r performance_measures_print, echo=FALSE}
pm_taxon
```

The marginal effects of an ENM, in the form of Partial Dependency Profile (PDP) and Accumulated Local Effect (ALE) plots (Molnar, 2018) can also be viewed using the `elements::plot_me` function.

```{r me_plot_create, include=FALSE}
me_plot <- elements::plot_me(taxon = "stellaria_graminea", 
                             me_type = "ale", 
                             free_y = FALSE, 
                             presences = TRUE,
                             vars = c("L", "M", "N", "R", "S", "SD", "GP", "bio05", "bio06", "bio16", "bio17"))
```

```{r me_plot_code, eval=FALSE}
elements::plot_me(taxon = "stellaria_graminea", 
                  me_type = "ale", 
                  free_y = FALSE, 
                  presences = TRUE,
                  vars = c("L", "M", "N", "R", "S", "SD", "GP", "bio05", "bio06", "bio16", "bio17"))
```

```{r me_plot_print, echo=FALSE, fig.dpi=800, fig.height=7, fig.width=7}
me_plot
```

## References

Hutchinson, G.E., 1957. Concluding Remarks. Cold Spring Harbor Symposia on Quantitative Biology 22, 415–427. https://doi.org/10.1101/SQB.1957.022.01.039

Meyer, D., Dimitriadou, E., Hornik, K., Weingessel, A., Leisch, F., 2024. e1071: Misc Functions of the Department of Statistics, Probability Theory Group (Formerly: E1071), TU Wien. https://doi.org/10.32614/CRAN.package.e1071

Molnar, C., 2018. iml: An R package for Interpretable Machine Learning. JOSS 3, 786. https://doi.org/10.21105/joss.00786

Peng, R.D., 2005. filehash: Simple Key-Value Database. https://doi.org/10.32614/CRAN.package.filehash

Schratz, P., Becker, M., Lang, M., Brenning, A., 2024. mlr3spatiotempcv: Spatiotemporal Resampling Methods for Machine Learning in R. Journal of Statistical Software 111, 1–36. https://doi.org/10.18637/jss.v111.i07



