---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# elements

<!-- badges: start -->
[![Generic badge](https://img.shields.io/badge/Version-0.5.0-green.svg)]()
[![License: GPL v3.0](https://img.shields.io/badge/License-GPL v3.0-lightgrey.svg)](https://opensource.org/license/lgpl-3-0)
[![Lifecycle:Experimental](https://img.shields.io/badge/Lifecycle-Experimental-339999)]()
<!-- badges: end -->

`elements` is a R package containing correlative, realised Ecological Niche Models (ENMs) for the most prevalent vascular plants, bryophytes, and terricolous lichens in the European Vegetation Archive (EVA) (Chytrý et al., 2016).

The code used to produce the models is available here [https://github.com/NERC-CEH/elementsAnalysis](https://github.com/NERC-CEH/elementsAnalysis).

<details>
<summary><h2 style="display:inline-block">Methods</h2></summary>

The ENMs adhere to Hutchinsonian conceptualisation of the ecological niche as an n-dimensional hypervolume (Hutchinson, 1957). The environmental variables forming the dimensions of each hypervolume consist of:

* Seven unweighted, plot-mean Ecological Indicator Values (EIVs): 
  * **M** - Soil Moisture<sup>1</sup>
  * **R** - Reaction<sup>1</sup>
  * **N** - Soil Nitrogen<sup>1</sup>
  * **L** - Light<sup>1</sup>
  * **GP** - Grazing Pressure<sup>2</sup>
  * **SD** - Soil Disturbance<sup>2</sup>
  * **S** - Salinity<sup>3</sup>
* Four bioclimatic variables: 
  * **bio05** - Maximum temperature in the warmest month<sup>4</sup>
  * **bio06** - Minimum temperature in the coldest month<sup>4</sup>
  * **bio16** - Precipitation in the wettest quarter<sup>4</sup>
  * **bio17** - Precipitation in the driest quarter<sup>4</sup>
  
Support Vector Machine (SVM) models, which form a hyperplane between the presence and absence hypervolumes,
were trained and tested using using the `mlr3` ecosystem of R packages (Lang et al., 2019).
The raw `e1071` (Meyer et al., 2024) SVM models are bundled in `elements`.

For more information please see Marshall et al (in prep).

<hr width="100%" size="1">

<p style="font-size: small !important">
<sup>1</sup>(Dengler et al., 2023), <sup>2</sup>(Midolo et al., 2023), <sup>3</sup>(Tichy et al., 2023), <sup>4</sup>(Copernicus Climate Change Service, 2021)
</p>

</details>

<details>
<summary><h2 style="display:inline-block">Installation</h2></summary>

The Github repository containing the `elements` package [(https://github.com/NERC-CEH/elements)](https://github.com/NERC-CEH/elements) 
includes all files apart from the "./inst/extdata/Models" object as it is above the 100MB limit.
To retrieve the `elements` package including the "Models" object download the latest
version from Zenodo here - .

Alternatively, you can download the package using the `zen4R` R package from a particular DOI.

``` r
zen4R::download_zenodo("10.5281/zenodo.<notyetinzotero>", path = "path_to_file")
```

After retrieving the package to install `elements` run:

``` r
install.packages(file.path("path_to_file", "elements_0.5.0.tar.gz"), repos = NULL, type = "source")
```

Downloading and installing the package may take a few minutes.

Note: `elements` has two dependencies, `e1071` and `filehash`, which must also be installed.

</details>

<details>
<summary><h2 style="display:inline-block">Model Usage</h2></summary>

### Retrieving models

```{r load_packages, include = FALSE}
library(e1071)
library(elements)
library(filehash)
```

```{r load_packages_print, include = TRUE, eval=FALSE}
library(e1071)
library(elements)
library(filehash)
```

Due to the total size of the `r nrow(elements::TaxonomicBackbone)` ENMs currently included in `elements` (1.8GB when compressed, 7.5GB in memory)
the ENMs are not exported in a .rda object. Instead they are made available through a `filehash` (Peng, 2005) database, which provides access to the ENMs without loading all models into memory.
To access the ENMs a connection to this database must be initialised using `elements::startup`.

```{r retrieve_models}
elements::startup()

model <- Models[["stellaria_graminea"]]
```

```{r print_model, echo=FALSE}
print(model)
```

### Using the models

The raw ENMs retrieved using the method above can be used as regular `e1071` SVM model objects. Alternatively, the helper function `elements::predict_occ_taxon` retrieves a model using the method above, generates predictions, and formats the results as a data frame.

```{r use_predict_occ}
results <- elements::predict_occ_taxon(taxon = "stellaria_graminea", predictors = elements::ExampleData1, pa = "Present", limit = NULL, dp = 2, append_predictors = FALSE)
```

```{r print_predict_occ_results, echo=FALSE}
head(results)
```

An additional helper function `elements::predict_occ` can generate predictions for multiple taxa, by either specifing the taxa to model in the 'taxa_codes' argument, or by setting 'taxa_codes' to NULL and including an additional column in the predictors data frame named 'taxon_code'.

```{r use_predict_occ_dfr}
results <- elements::predict_occ(taxa_codes = NULL, predictors = elements::ExampleData2, pa = "Present", limit = NULL, dp = 2, append_predictors = FALSE)
```

```{r print_predict_occ_dfr_results, echo=FALSE}
head(results)
```

Two helper arguments provide additional functionality in controlling model use.
First, is the 'limit' argument, which assigns probability values of zero if one or more predictors are outside a specified range e.g. the 10% and 90% quantiles (see `elements::NicheWidths`).
Second, is the 'holdopt' argument, which holds specified variable values at their optima (as defined by the mean value present in `elements::NicheWidths`).

As a simple demonstration, below two sets of predictions for *Stellaria graminea* are generated, holding all variables apart from N at their optima: 1) with no limit set, and 2) with a limit set to the 1% and 99% quantiles.

```{r limit_and_holdopt_demo}
n_gradient <- data.frame("N" = seq(0, 10, 0.01))

vary_N_no_limit <- elements::predict_occ_taxon(taxon = "stellaria_graminea", predictors = n_gradient,
                                               pa = "Present", limit = NULL, holdopt = c("bio05", "bio06", "bio16", "bio17", "GP", "L", "M", "R", "S", "SD"),
                                               dp = 2, append_predictors = TRUE)

vary_N_limit <- elements::predict_occ_taxon(taxon = "stellaria_graminea", predictors = n_gradient,
                                            pa = "Present", limit = "q01_q99", holdopt = c("bio05", "bio06", "bio16", "bio17", "GP", "L", "M", "R", "S", "SD"),
                                            dp = 2, append_predictors = TRUE)
```

```{r limit_and_holdopt_demo_create_plots, include=FALSE}
plot.new()
graphics::plot(NULL, xlim = c(0, 10), ylim = c(0, 1), xlab = "N", ylab = "Probability", main = "No limitation", cex.lab = 0.75, cex.axis = 0.75, cex.main = 0.75)
graphics::lines(x = vary_N_no_limit$N, y = vary_N_no_limit$Present)
vary_N_no_limit_plot <- grDevices::recordPlot()

plot.new()
graphics::plot(NULL, xlim = c(0, 10), ylim = c(0, 1), xlab = "N", ylab = "Probability", main = "Limitation", cex.lab = 0.75, cex.axis = 0.75, cex.main = 0.75)
graphics::lines(x = vary_N_limit$N, y = vary_N_limit$Present)
vary_N_limit_plot <- grDevices::recordPlot()
```

```{r limit_and_holdopt_demo_plots, echo=FALSE, fig.dpi=800, fig.height=3.5, fig.width=3.5, out.width="50%"}
vary_N_no_limit_plot
vary_N_limit_plot
```

Please note that as ten out of the eleven variables are held at their optima the predicted probabilities will be high as the influence of unsuitable N values will be partially offset.
Consequently, the shape of the response curves above will be wider than the corresponding PDP plot produced with the `elements::plot_me` function (see the **Model Inspection** section below).

### Shutting down

At the end of the analyis run `elements::shutdown` to close the connection to the filehash database.

```{r shutdown}
elements::shutdown()
```

</details>

<details>
<summary><h2 style="display:inline-block">Model Inspection</h2></summary>

Several datasets are available to examine the ENM model performance and aid in model interpretation.

The performance measures can be retrieved from the `elements::PerformanceMeasures` object; 
for example, below the balanced accuracy from the random holdout sample and the model tuning spatio-temporal cross-validation (Schratz et al., 2024) are displayed.

```{r performance_measures}
pm_taxon <- subset(elements::PerformanceMeasures, taxon_code == "stellaria_graminea", select = c("Holdout.BalancedAccuracy", "STCV.BalancedAccuracy"))
```

```{r performance_measures_print, echo=FALSE}
pm_taxon
```

The marginal effects of an ENM, in the form of Partial Dependency Profile (PDP) and Accumulated Local Effect (ALE) plots (Molnar, 2018) can also be viewed using the `elements::plot_me` function. By setting the 'presences' argument to TRUE a box and whiskers plot showing the distribution of presences is overlaid and by setting the 'eivs' argument to TRUE a point and arrows showing the EIV and niche width values are overlaid, where available in `elements::VariableData`.

```{r me_plot_taxon_create, include=FALSE}
me_plot_taxon <- elements::plot_me(taxa = "stellaria_graminea", 
                                   me_type = "ale", 
                                   free_y = TRUE, 
                                   presences = TRUE,
                                   eivs = TRUE,
                                   vars = c("L", "M", "N", "R", "S", "SD", "GP", "bio05", "bio06", "bio16", "bio17"))
```

```{r me_plot_taxon_code, eval=FALSE}
elements::plot_me(taxa = "stellaria_graminea", 
                  me_type = "ale", 
                  free_y = TRUE, 
                  presences = TRUE,
                  eivs = TRUE,
                  vars = c("L", "M", "N", "R", "S", "SD", "GP", "bio05", "bio06", "bio16", "bio17"))
```

```{r me_plot_taxon_print, echo=FALSE, fig.dpi=800, fig.height=7, fig.width=7}
me_plot_taxon
```

Multiple taxa can be supplied in the "taxa" argument, in this case the ability to plot presence box and whisker plots and EIV points and arrows is disabled.

```{r me_plot_taxa_create, include=FALSE}
me_plot_taxa <- elements::plot_me(taxa = c("galium_boreale", "galium_sylvaticum", "galium_uliginosum"), 
                                  me_type = "pdp", 
                                  normalise = TRUE,
                                  vars = c("L", "M", "N", "R", "S", "SD", "GP", "bio05", "bio06", "bio16", "bio17"))
```

```{r me_plot_taxa_code, eval=FALSE}
elements::plot_me(taxa = c("galium_boreale", "galium_sylvaticum", "galium_uliginosum"), 
                  me_type = "pdp", 
                  normalise = TRUE,
                  vars = c("L", "M", "N", "R", "S", "SD", "GP", "bio05", "bio06", "bio16", "bio17"))
```

```{r me_plot_taxa_print, echo=FALSE, fig.dpi=800, fig.height=7, fig.width=7}
me_plot_taxa
```

<details>
<summary><h3 style="display:inline-block">&#9888; A note on ALE plots</h3></summary>

```{r get_gr_s_nw_max, include=FALSE}
nw <- elements::NicheWidths
nw_gr_s_max <- nw[nw[["taxon_code"]] == "gymnocarpium_robertianum" & nw[["variable"]] == "S", ][["max"]]
```

In some instances the ALE curves may appear 'inverted' and not ecologically realistic, this is most often seen in situations where the distribution of presences along a variable gradient is extremely narrow and/or where there is a non-unimodal distribution, which causes extrapolation issues in the ALE calculations. For example, *Gymnocarpium robertianum* has a extremely narrow distribution of plot-mean S values, with a maximum value of `r nw_gr_s_max`.

In these instances it is important to also visualise the PDP plots, which should then be prioritised when inspecting the shape of the univariate response.

```{r me_plot_gr_s_create, include=FALSE}
me_plot_gr_s_ale <- elements::plot_me(taxa = "gymnocarpium_robertianum", 
                                      me_type = "ale", 
                                      free_y = TRUE, 
                                      presences = TRUE,
                                      eivs = TRUE,
                                      vars = c("S"))

me_plot_gr_s_pdp <- elements::plot_me(taxa = "gymnocarpium_robertianum", 
                                      me_type = "pdp", 
                                      free_y = TRUE, 
                                      presences = TRUE,
                                      normalise = FALSE,
                                      eivs = TRUE,
                                      vars = c("S"))
```

```{r me_plot_gr_s_print, echo=FALSE, fig.dpi=800, fig.height=3.5, fig.width=3.5, out.width="50%"}
me_plot_gr_s_ale
me_plot_gr_s_pdp
```


</details>

</details>

<details>
<summary><h2 style="display:inline-block">References</h2></summary>

Chytrý, M., Hennekens, S.M., Jiménez-Alfaro, B., Knollová, I., Dengler, J., Jansen, F., Landucci, F., Schaminée, J.H.J., Aćić, S., Agrillo, E., Ambarlı, D., Angelini, P., Apostolova, I., Attorre, F., Berg, C., Bergmeier, E., Biurrun, I., Botta-Dukát, Z., Brisse, H., Campos, J.A., Carlón, L., Čarni, A., Casella, L., Csiky, J., Ćušterevska, R., Dajić Stevanović, Z., Danihelka, J., De Bie, E., de Ruffray, P., De Sanctis, M., Dickoré, W.B., Dimopoulos, P., Dubyna, D., Dziuba, T., Ejrnæs, R., Ermakov, N., Ewald, J., Fanelli, G., Fernández-González, F., FitzPatrick, Ú., Font, X., García-Mijangos, I., Gavilán, R.G., Golub, V., Guarino, R., Haveman, R., Indreica, A., Işık Gürsoy, D., Jandt, U., Janssen, J.A.M., Jiroušek, M., Kącki, Z., Kavgacı, A., Kleikamp, M., Kolomiychuk, V., Krstivojević Ćuk, M., Krstonošić, D., Kuzemko, A., Lenoir, J., Lysenko, T., Marcenò, C., Martynenko, V., Michalcová, D., Moeslund, J.E., Onyshchenko, V., Pedashenko, H., Pérez-Haase, A., Peterka, T., Prokhorov, V., Rašomavičius, V., Rodríguez-Rojo, M.P., Rodwell, J.S., Rogova, T., Ruprecht, E., Rūsiņa, S., Seidler, G., Šibík, J., Šilc, U., Škvorc, Ž., Sopotlieva, D., Stančić, Z., Svenning, J.-C., Swacha, G., Tsiripidis, I., Turtureanu, P.D., Uğurlu, E., Uogintas, D., Valachovič, M., Vashenyak, Y., Vassilev, K., Venanzoni, R., Virtanen, R., Weekes, L., Willner, W., Wohlgemuth, T., Yamalov, S., 2016. European Vegetation Archive (EVA): an integrated database of European vegetation plots. Applied Vegetation Science 19, 173–180. https://doi.org/10.1111/avsc.12191

Copernicus Climate Change Service, 2021. Downscaled bioclimatic indicators for selected regions from 1950 to 2100 derived from climate projections. https://doi.org/10.24381/CDS.0AB27596

Dengler, J., Jansen, F., Chusova, O., Hüllbusch, E., Nobis, M.P., Meerbeek, K.V., Axmanová, I., Bruun, H.H., Chytrý, M., Guarino, R., Karrer, G., Moeys, K., Raus, T., Steinbauer, M.J., Tichý, L., Tyler, T., Batsatsashvili, K., Bita-Nicolae, C., Didukh, Y., Diekmann, M., Englisch, T., Fernández-Pascual, E., Frank, D., Graf, U., Hájek, M., Jelaska, S.D., Jiménez-Alfaro, B., Julve, P., Nakhutsrishvili, G., Ozinga, W.A., Ruprecht, E.-K., Šilc, U., Theurillat, J.-P., Gillet, F., 2023. Ecological Indicator Values for Europe (EIVE) 1.0. Vegetation Classification and Survey 4, 7–29. https://doi.org/10.3897/VCS.98324

Hutchinson, G.E., 1957. Concluding Remarks. Cold Spring Harbor Symposia on Quantitative Biology 22, 415–427. https://doi.org/10.1101/SQB.1957.022.01.039

Lang, M., Binder, M., Richter, J., Schratz, P., Pfisterer, F., Coors, S., Au, Q., Casalicchio, G., Kotthoff, L., Bischl, B., 2019. mlr3: A modern object-oriented machine learning framework in R. Journal of Open Source Software 4, 1903. https://doi.org/10.21105/joss.01903

Meyer, D., Dimitriadou, E., Hornik, K., Weingessel, A., Leisch, F., 2024. e1071: Misc Functions of the Department of Statistics, Probability Theory Group (Formerly: E1071), TU Wien. https://doi.org/10.32614/CRAN.package.e1071

Midolo, G., Herben, T., Axmanová, I., Marcenò, C., Pätsch, R., Bruelheide, H., Karger, D.N., Aćić, S., Bergamini, A., Bergmeier, E., Biurrun, I., Bonari, G., Čarni, A., Chiarucci, A., De Sanctis, M., Demina, O., Dengler, J., Dziuba, T., Fanelli, G., Garbolino, E., Giusso del Galdo, G., Goral, F., Güler, B., Hinojos-Mendoza, G., Jansen, F., Jiménez-Alfaro, B., Lengyel, A., Lenoir, J., Pérez-Haase, A., Pielech, R., Prokhorov, V., Rašomavičius, V., Ruprecht, E., Rūsiņa, S., Šilc, U., Škvorc, Ž., Stančić, Z., Tatarenko, I., Chytrý, M., 2023. Disturbance indicator values for European plants. Global Ecology and Biogeography 32, 24–34. https://doi.org/10.1111/geb.13603

Molnar, C., 2018. iml: An R package for Interpretable Machine Learning. Journal of Open Source Software 3, 786. https://doi.org/10.21105/joss.00786

Peng, R.D., 2005. filehash: Simple Key-Value Database. https://doi.org/10.32614/CRAN.package.filehash

Schratz, P., Becker, M., Lang, M., Brenning, A., 2024. mlr3spatiotempcv: Spatiotemporal Resampling Methods for Machine Learning in R. Journal of Statistical Software 111, 1–36. https://doi.org/10.18637/jss.v111.i07

Tichý, L., Axmanová, I., Dengler, J., Guarino, R., Jansen, F., Midolo, G., Nobis, M.P., Van Meerbeek, K., Aćić, S., Attorre, F., Bergmeier, E., Biurrun, I., Bonari, G., Bruelheide, H., Campos, J.A., Čarni, A., Chiarucci, A., Ćuk, M., Ćušterevska, R., Didukh, Y., Dítě, D., Dítě, Z., Dziuba, T., Fanelli, G., Fernández-Pascual, E., Garbolino, E., Gavilán, R.G., Gégout, J.-C., Graf, U., Güler, B., Hájek, M., Hennekens, S.M., Jandt, U., Jašková, A., Jiménez-Alfaro, B., Julve, P., Kambach, S., Karger, D.N., Karrer, G., Kavgacı, A., Knollová, I., Kuzemko, A., Küzmič, F., Landucci, F., Lengyel, A., Lenoir, J., Marcenò, C., Moeslund, J.E., Novák, P., Pérez-Haase, A., Peterka, T., Pielech, R., Pignatti, A., Rašomavičius, V., Rūsiņa, S., Saatkamp, A., Šilc, U., Škvorc, Ž., Theurillat, J.-P., Wohlgemuth, T., Chytrý, M., 2023. Ellenberg-type indicator values for European vascular plant species. Journal of Vegetation Science 34, e13168. https://doi.org/10.1111/jvs.13168

</details>



