---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

```{r load_packages, include = FALSE}
library(e1071)
library(elements)
library(filehash)
```

# elements

<!-- badges: start -->
[![Generic badge](https://img.shields.io/badge/Version-0.0.1-green.svg)]()
[![License: CC BY 4.0](https://img.shields.io/badge/License-LGPL 3.0-lightgrey.svg)](https://opensource.org/license/lgpl-3-0)
[![Lifecycle:Experimental](https://img.shields.io/badge/Lifecycle-Experimental-339999)]()
[![R-CMD-check](https://github.com/NERC-CEH/elements/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/NERC-CEH/elements/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`elements` is a R package containing **e**co**l**ogical nich**e** **m**od**e**ls for the pla**n**t**s** of Europe. More specifically, elements contains Ecological Niche Models (ENMs) for the most prevalent vascular plants, bryophytes, and terricolous lichens in the European Vegetation Archive.

The ENMs adhere to Hutchinsonian conceptualisation of the ecological niche as an n-dimensional hypervolume (Hutchinson, 1957) and are comprised of `e1071` (Meyer et al., 2024) Support Vector Machine (SVM) models.

For more information please see Marshall et al (in prep).

The code used to produce the models is available here [https://github.com/NERC-CEH/elementsAnalysis](https://github.com/NERC-CEH/elementsAnalysis).

<details>
<summary><h2 style="display:inline-block">Installation</h2></summary>

You can install the development version of elements from Github with:

``` r
if(!require("remotes")) install.packages("remotes")
remotes::install_github("NERC-CEH/elements")
```

Note: `elements` has two dependencies, `e1071` and `filehash`, which must also be installed.

</details>

<details>
<summary><h2 style="display:inline-block">Usage</h2></summary>

### Retrieving models

Due to the total size of the `r length(elements::ModellingSpecies)` ENMs currently included in `elements`
the ENMs are not exported in a .rda object. Instead they are made available through a `filehash` (Peng, 2005) database, which provides access to the ENMs without loading all models into memory.
To access the ENMs a connection to this database must be initialised using `elements::startup`.

```{r retrieve_models}
elements::startup()

model <- OccModels[["stellaria_graminea"]]
```

```{r print_model, echo=FALSE}
print(model)
```

### Using the models

The raw ENMs retrieved using the method above can be used as regular `e1071` SVM model objects. Alternatively, the helper function `elements::predict_occ_taxon` retrieves a model using the method above, generates predictions, and formats the results as a data frame.

```{r use_predict_occ}
results <- elements::predict_occ_taxon(taxon = "stellaria_graminea", predictors = elements::ExampleData1, pa = "Present", dp = 2, append_predictors = FALSE)
```

```{r print_predict_occ_results, echo=FALSE}
head(results)
```

An additional helper function `elements::predict_occ` can generate predictions for multiple taxa, by either specifing the taxa to model in the 'taxa_codes' argument, or by setting 'taxa_codes' to NULL and including an additional column in the predictors data frame named 'taxon_code'.

```{r use_predict_occ_dfr_1}
results <- elements::predict_occ(taxa_codes = NULL, predictors = elements::ExampleData2, pa = "Present", dp = 2, append_predictors = FALSE)
```

```{r print_predict_occ_dfr_1_results, echo=FALSE}
head(results)
```

```{r use_predict_occ_dfr_2}
results <- elements::predict_occ(taxa_codes = c("stellaria_graminea", "silene_flos-cuculi"), predictors = elements::ExampleData1, pa = "Present", dp = 2, append_predictors = FALSE)
```

```{r print_predict_occ_dfr_2_results, echo=FALSE}
head(results)
```

### Shutting down

At the end of the analyis run `elements::shutdown` to close the connection to the filehash database.

```{r shutdown}
elements::shutdown()
```

</details>

<details>
<summary><h2 style="display:inline-block">Inspecting Models</h2></summary>

Several datasets are available to examine the ENM model performance and aid in model interpretation.

The performance measures can be retrieved from the `elements::PerformanceMeasures` object; 
for example, below the balanced accuracy from the random holdout sample and the model tuning spatio-temporal cross-validation (Schratz et al., 2024) are displayed.

```{r performance_measures}
pm <- elements::PerformanceMeasures

pm_taxon <- pm[pm[["taxon_code"]] == "stellaria_graminea", c(1, 7, 8)]
```

```{r performance_measures_print, echo=FALSE}
pm_taxon
```

The marginal effects of an ENM, in the form of Partial Dependency Profile (PDP) and Accumulated Local Effect (ALE) plots (Molnar, 2018) can also be viewed using the `elements::plot_me` function.

```{r me_plot_create, include=FALSE}
me_plot <- elements::plot_me(taxon = "stellaria_graminea", 
                             me_type = "ale", 
                             free_y = FALSE, 
                             presences = TRUE,
                             vars = c("L", "M", "N", "R", "S", "SD", "GP", "bio05", "bio06", "bio16", "bio17"))
```

```{r me_plot_code, eval=FALSE}
elements::plot_me(taxon = "stellaria_graminea", 
                  me_type = "ale", 
                  free_y = FALSE, 
                  presences = TRUE,
                  vars = c("L", "M", "N", "R", "S", "SD", "GP", "bio05", "bio06", "bio16", "bio17"))
```

```{r me_plot_print, echo=FALSE, fig.dpi=800, fig.height=7, fig.width=7}
me_plot
```

</details>

## References

Hutchinson, G.E., 1957. Concluding Remarks. Cold Spring Harbor Symposia on Quantitative Biology 22, 415–427. https://doi.org/10.1101/SQB.1957.022.01.039

Meyer, D., Dimitriadou, E., Hornik, K., Weingessel, A., Leisch, F., 2024. e1071: Misc Functions of the Department of Statistics, Probability Theory Group (Formerly: E1071), TU Wien. https://doi.org/10.32614/CRAN.package.e1071

Molnar, C., 2018. iml: An R package for Interpretable Machine Learning. JOSS 3, 786. https://doi.org/10.21105/joss.00786

Peng, R.D., 2005. filehash: Simple Key-Value Database. https://doi.org/10.32614/CRAN.package.filehash

Schratz, P., Becker, M., Lang, M., Brenning, A., 2024. mlr3spatiotempcv: Spatiotemporal Resampling Methods for Machine Learning in R. Journal of Statistical Software 111, 1–36. https://doi.org/10.18637/jss.v111.i07



