---
title: "Scenario Development"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r load_packages, include = FALSE}
library(e1071)
library(elements)
library(filehash)
```

```{r load_packages_print, include = TRUE, eval=FALSE}
library(e1071)
library(elements)
library(filehash)
```

## Scenarios

`elements` is designed to be used to model scenarios of environmental change considering multiple interacting drivers. The object `elements::ExampleScenarios` provides a basic set of example scenarios
for the RF1 plot in `elements::ExamplePlots`: 
(A) Climate Change: a.i - ssp126, a.ii - ssp245, a.iii - ssp370, a.iv - ssp585.
(B) Grazing: b.i - Grazing intensification, b.i - Grazing reduction
(C) Nutrient: c.i - Nutrient Enrichment, c.ii - Nutrient Reduction
(D) Interactions: d.i - Climate Change + Nutrient Reduction + Grazing Reduction, 
d.ii - Climate Change + Nutrient Addition + Grazing Reduction, 
d.iii - Climate Change + Nutrient Reduction + Grazing Intensification, 
d.iv - Climate Change + Nutrient Addition + Grazing Intensification
Below the predicted probabilities for the taxa in plot RF1 in `elements::ExamplePlots` for all scenarios are displayed.

```{r scenario_results}
elements::startup()

rf1_taxa <- elements::ExamplePlots |>
  dplyr::filter(plot_id == "RF1") |>
  dplyr::pull(taxon_code)

scenario_results <- elements::predict_occ(taxa = rf1_taxa,
                                          predictors = elements::ExampleScenarios,
                                          append = "ids") |>
  dplyr::left_join(elements::ModellingTaxaLookup |> dplyr::distinct(taxon_name, taxon_code), by = "taxon_code")
```

```{r create_scenario_example_plot, include = FALSE}
scenarios_len <- length(unique(scenario_results$taxon_name))
scenario_col_pal <- sample(paletteer::paletteer_d("khroma::smoothrainbow") |> as.character(), scenarios_len)
names(scenario_col_pal) <- unique(scenario_results$taxon_name)

scenario_full_names <- elements::ExampleScenarios |>
  dplyr::distinct(scenario_code, scenario_name) |>
  tibble::deframe()
  
scenario_example_plot <- ggplot2::ggplot() +
  ggplot2::geom_line(data = scenario_results |> dplyr::filter(scenario_stage != "baseline"), 
                     mapping = ggplot2::aes(x = timeslice, y = Present, color = taxon_name, group = taxon_name), 
                     position = ggplot2::position_dodge(width = 0.1)) +
  ggplot2::geom_point(data = scenario_results |> dplyr::filter(scenario_stage != "baseline"), 
                      mapping = ggplot2::aes(x = timeslice, y = Present, color = taxon_name, group = taxon_name), 
                      position = ggplot2::position_dodge(width = 0.1)) +
  ggplot2::geom_point(data = scenario_results |> 
                        dplyr::filter(scenario_stage == "baseline") |> 
                        dplyr::distinct(taxon_name, timeslice, Present), 
                      mapping = ggplot2::aes(x = timeslice, y = Present, color = taxon_name), 
                      position = ggplot2::position_dodge(width = 0.1)) +
  ggplot2::facet_wrap(dplyr::vars(scenario_code), 
                      ncol = 2,
                      labeller = ggplot2::as_labeller(scenario_full_names),
                      scales = "free_x") + 
  ggplot2::expand_limits(y = 0) +
  ggplot2::theme_classic() +
  ggplot2::theme(strip.background = ggplot2::element_blank(),
                 legend.key.spacing.y = ggplot2::unit(0.05, "cm"),
                 legend.key.height = ggplot2::unit(0.3, "cm"),
                 legend.position = "bottom") +
  ggplot2::scale_color_manual(values = scenario_col_pal) +
  ggplot2::guides(color = ggplot2::guide_legend(title = "Taxon", ncol = 2)) +
  ggplot2::ylab(NULL)
```

```{r scenario_example_plot, echo=FALSE, dpi = 800, fig.height=10, fig.width=7, out.width="100%", fig.align='center'}
scenario_example_plot
```

## Environmental Filtering

In addition to modelling the response of individual taxa over time let's inspect how the constant taxa in EUNIS habitats respond to scenarios in `elements::ExampleScenarios`,
by presenting the mean of the probabilities of each EUNIS habitats constant taxa, for the top EUNIS habiats.

```{r calc_env_filter_eunis}
eunis_taxa_codes <- elements::EUNISDiagnosticTaxa |> # elements::EUNISConstantTaxa
  tibble::as_tibble() |>
  dplyr::distinct(taxon_code) |>
  dplyr::mutate("taxon_match" = stringr::str_remove(string = taxon_code, pattern = "_aggr\\.$")) |>
  dplyr::rename("eunis_taxon_code" = "taxon_code")

elements_codes <- elements::ModellingTaxaLookup |>
  tibble::as_tibble() |>
  dplyr::distinct(taxon_code) |>
  dplyr::mutate("taxon_match" = stringr::str_remove(string = taxon_code, pattern = "_aggr$"))

elements_eunis_models_tbl <- elements_codes |>
  dplyr::inner_join(eunis_taxa_codes, by = "taxon_match") 

elements_eunis_models <- elements_eunis_models_tbl |>
  dplyr::pull(taxon_code) |>
  setdiff("glyceria_maxima")

predictors_scenanrios <- elements::ExampleScenarios |> 
  dplyr::distinct()

eunis_filter_results <- elements::env_filter(predictors = predictors_scenanrios, 
                                             taxa = elements_eunis_models, 
                                             screen = FALSE,
                                             method = "svm", 
                                             limit = "min_max", 
                                             exclude = NULL, 
                                             threshold = NULL, 
                                             append = "ids")
```

```{r q_summed_probs_plot, echo=FALSE, include=FALSE}
eunis_filter_results_sum <- elements::EUNISConstantTaxa |>
  dplyr::filter(taxon_code %in% elements_eunis_models_tbl$eunis_taxon_code) |>
  dplyr::group_by(eunis_code) |>
  dplyr::mutate("n" = dplyr::n()) |>
  dplyr::ungroup() |>
  dplyr::inner_join(eunis_filter_results, by = "taxon_code", relationship = "many-to-many") |>
  dplyr::summarise("sum" = sum(Present, na.rm = TRUE),
                   .by = c("eunis_name", "eunis_code", "timeslice", "scenario_stage", "scenario_code", "n")) |>
  dplyr::mutate("sum" = sum / n)

top_eunis_names <- eunis_filter_results_sum |>
  dplyr::group_by(timeslice, scenario_stage, scenario_code) |>
  dplyr::slice_max(sum, n = 2) |>
  dplyr::ungroup() |>
  dplyr::distinct(eunis_name) |>
  dplyr::filter(eunis_name != c("Wet inland cliff")) |>
  dplyr::pull(eunis_name)
  
eunis_filter_results_sum_top <- eunis_filter_results_sum |>
  dplyr::filter(eunis_name %in% top_eunis_names)

eunis_len <- length(top_eunis_names)
eunis_col_pal <- unname(Polychrome::kelly.colors()[-1][1:eunis_len])
names(eunis_col_pal) <- top_eunis_names
```

The mean probability for the constant taxa in the top EUNIS habitat in the baseline (Q24 - Intermediate fen and soft-water spring mire)
decreases in all example scenarios.

```{r q_summed_probs_adj_plot, echo=FALSE, include=FALSE}
eunis_filter_results_adj_plot <- ggplot2::ggplot() +
  ggplot2::geom_line(data = eunis_filter_results_sum_top |> dplyr::filter(scenario_stage != "baseline"), 
                     mapping = ggplot2::aes(x = timeslice, y = sum, color = eunis_name, group = eunis_name), 
                     position = ggplot2::position_dodge(width = 0.1)) +
  ggplot2::geom_point(data = eunis_filter_results_sum_top |> dplyr::filter(scenario_stage != "baseline"), 
                      mapping = ggplot2::aes(x = timeslice, y = sum, color = eunis_name, group = eunis_name), 
                      position = ggplot2::position_dodge(width = 0.1)) +
  ggplot2::geom_point(data = eunis_filter_results_sum_top |>
                        dplyr::filter(scenario_stage == "baseline") |>
                        dplyr::distinct(eunis_name, timeslice, sum),
                      mapping = ggplot2::aes(x = timeslice, y = sum, color = eunis_name),
                      position = ggplot2::position_dodge(width = 0.1)) +
  ggplot2::expand_limits(y = 0) +
  ggplot2::facet_wrap(dplyr::vars(scenario_code), 
                      ncol = 2,
                      labeller = ggplot2::as_labeller(scenario_full_names),
                      scales = "free_x") + 
  ggplot2::theme_classic() +
  ggplot2::theme(strip.background = ggplot2::element_blank(),
                 legend.key.spacing.y = ggplot2::unit(0.05, "cm"),
                 legend.key.height = ggplot2::unit(0.3, "cm"),
                 legend.position = "bottom") +
  ggplot2::scale_color_manual(values = eunis_col_pal) +
  ggplot2::guides(color = ggplot2::guide_legend(title = "EUNIS", ncol = 2)) +
  ggplot2::ylab(NULL)
```

```{r scenario_eunisQ_weighted_plot, echo=FALSE, dpi = 800, fig.height=10, fig.width=7, out.width="100%", fig.align='center'}
eunis_filter_results_adj_plot
```

```{r shutdown}
elements::shutdown()
```
