---
title: "Scenario Development"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r load_packages, include = FALSE}
library(e1071)
library(elements)
library(filehash)
```

```{r load_packages_print, include = TRUE, eval=FALSE}
library(e1071)
library(elements)
library(filehash)
```


## Scenarios

`elements` is designed to be used to model scenarios of environmental change considering multiple interacting drivers. The object `elements::ExampleScenarios` provides a basic set of example scenarios: (A) Climate Change - RCP4.5, (A) Climate Change - RCP8.5, (B) Grazing Intensification (+0.025GP per year), (B) Grazing Reduction (-0.025GP per year), and (C) Nutrient Enrichment (+0.25N per year) along with a Baseline scenario.
Below the predicted probabilities for the taxa in `elements::ExamplePlot` for Scenario C are displayed.

```{r scenario_c_results}
elements::startup()

scenario_results <- elements::predict_occ(taxa = elements::ExamplePlot$taxon_code,
                                          predictors = elements::ExampleScenarios,
                                          append = "ids")
```

```{r create_scenario_example_plot, include = FALSE}
scenarios_len <- length(unique(scenario_results$taxon_code))
scenario_col_pal <- ggsci::pal_uchicago("default")(scenarios_len)
names(scenario_col_pal) <- unique(scenario_results$taxon_code)
  
scenario_example_plot <- ggplot2::ggplot() +
  ggplot2::geom_line(data = scenario_results |> dplyr::filter(scenario != "Baseline"), 
                     mapping = ggplot2::aes(x = timeslice, y = Present, color = taxon_code, group = taxon_code), 
                     position = ggplot2::position_dodge(width = 0.1)) +
  ggplot2::geom_point(data = scenario_results |> dplyr::filter(scenario != "Baseline"), 
                      mapping = ggplot2::aes(x = timeslice, y = Present, color = taxon_code, group = taxon_code), 
                      position = ggplot2::position_dodge(width = 0.1)) +
  ggplot2::geom_point(data = scenario_results |> 
                        dplyr::filter(scenario == "Baseline") |> 
                        dplyr::distinct(taxon_code, timeslice, Present), 
                      mapping = ggplot2::aes(x = timeslice, y = Present, color = taxon_code), 
                      position = ggplot2::position_dodge(width = 0.1)) +
  ggplot2::facet_wrap(dplyr::vars(scenario), ncol = 2,, scales = "free_x") + 
  ggplot2::theme_classic() +
  ggplot2::theme(strip.background = ggplot2::element_blank(),
                 legend.position = "top") +
  ggplot2::scale_color_manual(values = scenario_col_pal) +
  ggplot2::guides(color = ggplot2::guide_legend(title = "Taxon", ncol = 2)) +
  ggplot2::ylab(NULL)
```

```{r scenario_example_plot, echo=FALSE, dpi = 800, fig.height=10, fig.width=7, out.width="100%", fig.align='center'}
scenario_example_plot
```

## Environmental Filtering

In addition to modelling the response of individual taxa over time let's inspect how the constant taxa in EUNIS habitats respond to scenarios in `elements::ExampleScenarios`,
by presenting the mean probability of each EUNIS habitats constant taxa.

```{r calc_env_filter_eunis}
eunis_taxa_models <- intersect(unique(elements::EUNISConstantTaxa$taxon_code), 
                               elements::TaxonomicBackbone$taxon_code)

predictors_scenanrios <- elements::ExampleScenarios |> 
  dplyr::select(-scenario_code) |>
  dplyr::distinct()

eunis_filter_results <- elements::env_filter(predictors = predictors_scenanrios, 
                                             taxa = eunis_taxa_models, 
                                             screen = TRUE, method = "svm", 
                                             limit = "min_max", 
                                             exclude = NULL, threshold = NULL, 
                                             append = "ids")
```

```{r q_summed_probs_plot, echo=FALSE, include=FALSE}
eunis_filter_results_sum <- elements::EUNISConstantTaxa |>
  dplyr::filter(taxon_code %in% eunis_taxa_models) |>
  dplyr::group_by(eunis_code) |>
  dplyr::mutate("n" = dplyr::n()) |>
  dplyr::ungroup() |>
  dplyr::inner_join(eunis_filter_results, by = "taxon_code", relationship = "many-to-many") |>
  dplyr::summarise("sum" = sum(Present, na.rm = TRUE),
                   .by = c("eunis_name", "eunis_code", "timeslice", "scenario", "n")) |>
  dplyr::mutate("sum" = sum / n)

top_eunis_names <- eunis_filter_results_sum |>
  dplyr::group_by(timeslice, scenario) |>
  dplyr::slice_max(sum, n = 1) |>
  dplyr::ungroup() |>
  dplyr::distinct(eunis_name) |>
  dplyr::pull(eunis_name)
  
eunis_filter_results_sum_top <- eunis_filter_results_sum |>
  dplyr::filter(eunis_name %in% top_eunis_names)

eunis_len <- length(top_eunis_names)
eunis_col_pal <- unname(Polychrome::kelly.colors()[-1][1:eunis_len])
names(eunis_col_pal) <- top_eunis_names
```

The mean probability for the constant taxa in top EUNIS wetland habitat in the baseline (Q24 - Intermediate fen and soft-water spring mire and S92 - Salix fen scrub)
decreases in all example scenarios.

```{r q_summed_probs_adj_plot, echo=FALSE, include=FALSE}
eunis_filter_results_adj_plot <- ggplot2::ggplot() +
  ggplot2::geom_line(data = eunis_filter_results_sum_top |> dplyr::filter(scenario != "Baseline"), 
                     mapping = ggplot2::aes(x = timeslice, y = sum, color = eunis_name, group = eunis_name), 
                     position = ggplot2::position_dodge(width = 0.1)) +
  ggplot2::geom_point(data = eunis_filter_results_sum_top |> dplyr::filter(scenario != "Baseline"), 
                      mapping = ggplot2::aes(x = timeslice, y = sum, color = eunis_name, group = eunis_name), 
                      position = ggplot2::position_dodge(width = 0.1)) +
  ggplot2::geom_point(data = eunis_filter_results_sum_top |>
                        dplyr::filter(scenario == "Baseline") |>
                        dplyr::distinct(eunis_name, timeslice, sum),
                      mapping = ggplot2::aes(x = timeslice, y = sum, color = eunis_name),
                      position = ggplot2::position_dodge(width = 0.1)) +
  ggplot2::ylim(c(0, 1)) +
  ggplot2::facet_wrap(dplyr::vars(scenario), ncol = 2,, scales = "free_x") + 
  ggplot2::theme_classic() +
  ggplot2::theme(strip.background = ggplot2::element_blank(),
                 legend.position = "top") +
  ggplot2::scale_color_manual(values = eunis_col_pal) +
  ggplot2::guides(color = ggplot2::guide_legend(title = "EUNIS", ncol = 2)) +
  ggplot2::ylab(NULL)
```

```{r scenario_eunisQ_weighted_plot, echo=FALSE, dpi = 800, fig.height=10, fig.width=7, out.width="100%", fig.align='center'}
eunis_filter_results_adj_plot
```

```{r shutdown}
elements::shutdown()
```